ChemKED: a human- and machine-readable data standard for chemical kinetics experiments

### 2017-07-18 pyked.tex
style name -> ijck
replace inline code highlighting with listings

% !TEX option = -shell-escape
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This work is licensed under the Creative Commons Attribution 4.0
% International License. To view a copy of this license, visit
% http://creativecommons.org/licenses/by/4.0/.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[12pt]{icck}

@@

This section will provide a brief introduction to the \ck{} database format.
This is not a complete listing of available fields and more complete examples of
\ck{} files can be found in the \pybox{tests} directory of the \pk{} package. In
addition, in Section~\ref{sec:usage-example}, we will provide examples of \ck{}
files. We indicate \yabox{yaml} keywords or values using text with gray background,
and \pybox{Python} code using text with a yellow background.

@@

The \yabox{value} can be a string, integer, or floating point number. The \ck{}
format is designed to include all of the information necessary to simulate a
given experiment and can be broken into two main sections. The first section
details all of the ``meta'' information about the experiment and the \ck{} file.
In this section, the common \yabox{author} mapping is used for any fields where
an author must be specified. The \yabox{author} mapping contains the following
fields:
%
\begin{itemize}
    \item \yabox{name} (required, string): The author's full name
    \item \yabox{ORCID} (optional, string): the author's ORCID number
\end{itemize}
%
More generally for the meta information, the required keys in this section
are:
%
\begin{itemize}
    \item \yabox{file-author} (required, \yabox{author}-type mapping): The
    author of the \ck{} file
    \item \yabox{file-version} (required, integer): The version of the \ck{}
    file
    \item \yabox{chemked-version} (required, string): The version of the \ck{}
    schema that this file conforms to
    \item \yabox{reference} (required, mapping): The reference information for
    the article associated with the data in the file. The fields in this
    mapping are:
    \begin{itemize}
        \item \yabox{authors} (required, sequence): A sequence of \yabox{author}
        mappings
        \item \yabox{journal} (required, string): The name of the publishing
        journal
        \item \yabox{year} (required, integer): The year of publication
        \item \yabox{doi} (optional, string): The article DOI
        \item \yabox{volume} (required, integer): The journal volume number
        \item \yabox{pages} (optional, string): The article pages
        \item \yabox{detail} (optional, string): A description of from where the
        data originated (e.g., figure or table number)
    \end{itemize}
    \item \yabox{experiment-type} (required, string): Currently, only
    \yabox{ignition delay} is supported
    \item \yabox{apparatus} (required, string): The type of apparatus used to
    perform the experiment (\yabox{shock tube} or \yabox{rapid compression machine})
\end{itemize}

The second section of the file encodes the experimental data. In the \ck{} file,
this is done in a sequence of mappings, with the following information required
in each element of the sequence:
%
\begin{itemize}
    \item \yabox{temperature} (required, sequence): The temperature of the
    experiment, with units
    \item \yabox{ignition-delay} (required, sequence): The ignition delay of the
    experiment, with units
    \item \yabox{pressure} (required, string): The pressure of the experiment,
    with units
    \item \yabox{composition} (required, sequence): The composition of the
    mixture in the experiment, a sequence of mappings with the following fields
    \begin{itemize}
        \item \yabox{species-name} (required, string): The name of the species
        \item \yabox{mole-fraction}, \yabox{mass-fraction}, \yabox{mole-percent}
        (required, float): The mole fraction or percent, or the mass fraction.
        Only one of the three is permitted in a given \yabox{composition}
        sequence.
        \item \yabox{InChI}, \yabox{SMILES}, \yabox{atomic-composition}, or
        \yabox{elemental-composition} (required, string or sequence): The InChI
        or SMILES representing the molecule, or its atomic or elemental
        composition (\yabox{atomic-composition} and
        \yabox{elemental-composition} are synonyms) as a sequence
    \end{itemize}
    \item \yabox{ignition-type} (required, mapping): The type of ignition
    measured in the experiments. The required fields are
    \begin{itemize}
        \item \yabox{type} (required, string): How the ignition delay was
        measured; one of \yabox{d/dt max} to indicate the ignition point is
        found at the maximum of the time derivative of the \yabox{target},
        \yabox{max} or \yabox{min} to indicate the ignition point is at the
        maximum or minimum of the \yabox{target}, or \yabox{1/2 max} to
        indicate the half-maximum point
        \item \yabox{target} (required, string): The target for the type of the
        ignition point; one of \yabox{temperature}, \yabox{pressure},
        \yabox{OH}, \yabox{OH*}, \yabox{CH}, or \yabox{CH*}
    \end{itemize}
\end{itemize}

The top-level key for the sequence is called \yabox{datapoints}. Each of the
quantities in an element must be specified with units. This should be done as a
single string associated with the first element of the sequence. The units of
the quantity are validated to ensure they have the appropriate dimensionality
for the quantity. In addition, each of the quantities in an element can
optionally be assigned an uncertainty. This uncertainty can be either absolute
or relative, and is specified as elements of the sequence of the associated key.
For example, the absolute uncertainty of the temperature and the relative
uncertainty of the ignition delay might be specified as follows:

@@

The basic user interface to the \ck{} file in \pk{} is through the
\pybox{ChemKED} class. The \pybox{ChemKED} class constructor takes the filename
of a \ck{} file or a Python dictionary containing the contents of a \ck{} file
as its argument. When the file or dictionary is loaded, it is validated to have
the correct format by the Python package Cerberus \autocite{cerberus}. The
schema used for validation of the \ck{} files is available publicly in the
source code repository for
PyKED\footnote{\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/chemked_schema.yaml}}.

The fields of the \ck{} file are stored as instance attributes of the
\pybox{ChemKED} class. The following attributes are available:
%
\begin{itemize}
    \item \pybox{chemked_version}, \pybox{file_version}, \pybox{file_author},
    \pybox{experiment_type}: Stores the values from the \ck{} file
    \item \pybox{reference}: Instance of a \pybox{namedtuple} containing all
    the information from the literature reference associated with the data
    \item \pybox{apparatus}: Instance of a \pybox{namedtuple} containing all
    the information about the apparatus used to perform the experiment
    \item \pybox{datapoints}: A Python list of \pybox{DataPoint} instances
\end{itemize}

The \pybox{DataPoint} class stores the information associated with a single data
point in the \ck{} file (i.e., a single element of the \yabox{datapoints}
sequence). Similar to the \pybox{ChemKED} class, information in the
\pybox{DataPoint} is stored as instance attributes:
%
\begin{itemize}
    \item \pybox{equivalence_ratio}: The value of the equivalence ratio, if
    present. For informational purposes only, no validation is done of the
    value.
    \item \pybox{ignition_delay}, \pybox{temperature}, \pybox{pressure},
    \pybox{pressure_rise}: The value of these data are stored as instances of
    the \pybox{Quantity} class from the Pint \autocite{Grecco2016} package,
    possibly with associated uncertainty
    \item \pybox{composition}: A list of dictionaries of the species and their
    respective amounts. The values are validated so that \yabox{mole-percent},
    \yabox{mole-fraction}, or \yabox{mass-fraction} cannot be mixed for a single
    data point, and so that the sum of the values is approximately 1.0, or
    100.0 for \yabox{mole-percent}.
    \item \pybox{composition_type}: A string indicating the type of composition
    information for the data point---one of \pybox{'mole-percent'},
    \pybox{'mole-fraction'}, or \pybox{'mass-fraction'}
    \item \pybox{volume_history}: If the \yabox{volume-history} of an RCM
    experiment is provided in the \ck{} file, it is stored in this attribute as
    a \pybox{namedtuple}, and the actual values are stored in NumPy arrays
    \autocite{vanderWalt:2011np}
    \item \pybox{compression_time}: If the \ck{} file encodes an RCM experiment,
    the length of the compression stroke may be included
    \item \pybox{ignition_type}: A dictionary specifying the method of the
    measurement of ignition delay
\end{itemize}

The \pybox{DataPoint} class defines two instance methods:
\pybox{get_cantera_mole_fraction()} and \pybox{get_cantera_mass_fraction()}.
With these methods, the composition of the reactant mixture can be output to a
format that can be used to set the composition of a
Cantera~\pybox{Solution}~\autocite{Cantera:2.3.0}.
The \pybox{composition} specification does not contain the molecular
weights of the components, so conversion between mole fractions and mass
fractions is not currently possible.

The \pybox{ChemKED} class defines one instance method:
\pybox{get_dataframe()}. Using this method, the user can get an instance of a
Pandas \pybox{DataFrame}~\autocite{pandas} that contains the information in the
list of \pybox{DataPoint}s. The user can specify the columns included in the
\pybox{DataFrame} by passing a list of column names to the
\pybox{output_columns} argument of the \pybox{get_dataframe()} method. The
possible columns are not case-sensitive and are:

\noindent\begin{itemize*}
    \item \pybox{'Temperature'}
    \item \pybox{'Pressure'}
    \item \pybox{'Ignition Delay'}
    \item \pybox{'Composition'}
    \item \pybox{'Equivalence Ratio'}
    \item \pybox{'Reference'}
    \item \pybox{'Apparatus'}
    \item \pybox{'Experiment Type'}
    \item \pybox{'File Author'}
    \item \pybox{'File Version'}
    \item \pybox{'ChemKED Version'}
\end{itemize*}

In addition, specific fields from the \pybox{Reference} and \pybox{Apparatus}
attributes can be included by specifying the name after a colon. These options
are:

\noindent\begin{itemize*}
    \item \pybox{'Reference:Volume'}
    \item \pybox{'Reference:Journal'}
    \item \pybox{'Reference:DOI'}
    \item \pybox{'Reference:Authors'}
    \item \pybox{'Reference:Detail'}
    \item \pybox{'Reference:Year'}
    \item \pybox{'Reference:Pages'}
    \item \pybox{'Apparatus:Kind'}
\end{itemize*}

% This weirdness is necessary to prevent the list from running over the margin
\noindent\begin{itemize*}
    \item \pybox{'Apparatus:Facility'}
    \item \pybox{'Apparatus:Institution'}
\end{itemize*}

Only the first author is printed when \pybox{Reference} or
\pybox{Reference:Authors} is selected because the whole author list may be
quite long.

@@

\pk{} is licensed under the permissive, open-source BSD 3-clause license. The
source code is publicly available on GitHub at
\url{https://github.com/pr-omethe-us/PyKED}~\autocite{PyKED}. Installation of stable versions
can be accomplished from a package repository, using either \pybox{pip} to install
from \href{https://pypi.python.org/pypi}{PyPI} (Python Package Index), or
\pybox{conda} to install from the
\href{https://anaconda.org/conda-forge}{\texttt{conda-forge}} channel. In
addition, \pk{} can be installed from the source code by running
\pybox|python setup.py install| or \pybox{pip install .} from the root
directory of the repository.

@@

The \ck{} file that will be used in this example can be found in the
\pybox{tests} directory of the PyKED
repository\footnote{\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/tests/testfile_rcm.yaml}}.
Examining that file, we find the first section specifies the information about
the \ck{} file itself:
%
\begin{yamlbox}
file-author:
  name: Kyle E Niemeyer
  ORCID: 0000-0003-4425-7097
file-version: (1, 0)
chemked-version: 0.0.1
\end{yamlbox}
%
Then, we find the information regarding the article in the literature from which
this data was taken. In this case, the dataset comes from the work of
\textcite{Mittal2006a}:
%
\begin{yamlbox}
reference:
  doi: 10.1002/kin.20180
  authors:
    - name: Gaurav Mittal
    - name: Chih-Jen Sung
      ORCID: 0000-0003-2046-8076
    - name: Richard A Yetter
  journal: International Journal of Chemical Kinetics
  year: 2006
  volume: 38
  pages: 516-529
  detail: Fig. 6, open circle
experiment-type: ignition delay
apparatus:
  kind: rapid compression machine
  institution: Case Western Reserve University
  facility: CWRU RCM
\end{yamlbox}
%
Finally, there is just a single datapoint in this file, which describes the ignition
delay for the experiment, the mixture composition, the initial temperature,
pressure, compression time, ignition type, and the volume history that specifies
how the volume of the reactor varies with time, for simulating the compression
stroke and post-compression processes.
%
\begin{yamlbox}
datapoints:
  - temperature: 297.4 kelvin
    ignition-delay: 1.0 ms
    pressure: 958.0 torr
    composition:
      - species-name: H2
        InChI: 1S/H2/h1H
        mole-fraction: 0.12500
      - species-name: O2
        InChI: 1S/O2/c1-2
        mole-fraction: 0.06250
      - species-name: N2
        InChI: 1S/N2/c1-2
        mole-fraction: 0.18125
      - species-name: Ar
        InChI: 1S/Ar
        mole-fraction: 0.63125
    ignition-type:
      target: pressure
      type: d/dt max
    compression-time: 38.0 ms
    volume-history:
      time:
        units: s
        column: 0
      volume:
        units: cm3
        column: 1
      values:
        - [0.00E+000, 5.47669375000E+002]
        - [1.00E-003, 5.46608789894E+002]
        - [2.00E-003, 5.43427034574E+002]
        - [3.00E-003, 5.38124109043E+002]
        ...
\end{yamlbox}
%
The values for the \yabox{volume-history} are truncated here to save space. A user
might want to load the information from this file to perform a simulation using
Cantera to calculate the ignition delay. All of the information required to
perform this simulation is present in the \ck{} file, with the exception of a
chemical kinetic model for \ce{H2}\slash {CO} combustion.
%
\begin{pythonbox}
import cantera as ct
from pyked import ChemKED

# Load the ChemKED file and retrieve the first element of the
# datapoints list, which is an instance of the DataPoint class
ck = ChemKED('testfile_rcm.yaml')
dp = ck.datapoints[0]
\end{pythonbox}
%
The initial temperature, pressure, and mixture composition can be read from the
instance of the \pybox{DataPoint} class. \pk{} uses Pint \pybox{Quantities} to
store values with units, while Cantera expects a floating point value in SI
units as input. Therefore, we use the built-in capabilities of Pint to convert
the units from those specified in the \ck{} file to SI units:
%
\begin{pythonbox}
T_initial = dp.temperature.to('K').magnitude
P_initial = dp.pressure.to('Pa').magnitude
X_initial = dp.get_cantera_mole_fraction()

@@

# Create the reactor and the outside environment
reac = ct.IdealGasReactor(gas)
env = ct.Reservoir(ct.Solution('air.xml'))
\end{pythonbox}
%
To apply the effect of the volume trace to the \pybox{IdealGasReactor}, a
\pybox{Wall} is installed between the reactor and the environment and assigned
a velocity. Although we do not show the details here, a reference implementation
of a class that computes a wall velocity given the volume history of the reactor
is available in CanSen~\autocite{cansen}, in the
\pybox{cansen.profile.VolumeProfile} class. Then, a Cantera \pybox{ReactorNet}
can be used to advance the state through autoignition, in this case to an end
time of \SI{50}{\milli\second}:
%
\begin{pythonbox}
time = dp.volume_history.time
volume = dp.volume_history.volume
ct.Wall(reac, env, velocity=VolumeProfile(time=time, volume=volume))

@@

# Integrate for 50 ms
while netw.time < 0.05:
    netw.step()
\end{pythonbox}
%
Although not shown in this example, the user would typically store information
about the \pybox{IdealGasReactor} (e.g., temperature, pressure, mass fractions)
at each time step of the integration for post-processing.

@@

The \ck{} file used in this example can be found in the \pybox{tests} directory of the \pk{}
repository\footnote{\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/tests/testfile_st_p5.yaml}}.
The data in this file comes from Stranic et al.~\autocite{Stranic:2012}, describing
shock-tube ignition delays for \textit{tert}-butanol. We have omitted the file
meta information below for space; the format is largely similar to the example in Section~\ref{sec:rcm-modeling}.
%
This \ck{} file specifies multiple data points with some common
conditions, including a common mixture composition and common definition of
ignition delay. Therefore, a \yabox{common-properties} section is specified.
%
\begin{yamlbox}
common-properties:
  composition: &comp
    - species-name: t-butanol
      InChI: 1S/C4H10O/c1-4(2,3)5/h5H,1-3H3
      mole-fraction: 0.003333333
    - species-name: O2
      InChI:  1S/O2/c1-2
      mole-fraction: 0.04
    - species-name: Ar
      InChI:  1S/Ar
      mole-fraction: 0.956666667
  ignition-type:  &ign
    target: OH* concentration
    type: 1/2 max
\end{yamlbox}
%
This block uses the ability of YAML files to define an anchor with the
\yabox{&} symbol and refer to that section later with the \yabox{*} symbol, as
can be seen in the definition of the \yabox{datapoints} section:
%
\begin{yamlbox}
datapoints:
  - temperature: 1459 kelvin
    ignition-delay: 347 us
    pressure: 1.60 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1389 kelvin
    ignition-delay: 756 us
    pressure: 1.67 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1497 kelvin
    ignition-delay: 212 us
    pressure: 1.55 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1562 kelvin
    ignition-delay: 105 us
    pressure: 1.50 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
\end{yamlbox}
The \yabox{common-properties} section is not required, but can save space and
help avoid errors when all data points share some common values.

>>>>

% !TEX option = -shell-escape
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This work is licensed under the Creative Commons Attribution 4.0
% International License. To view a copy of this license, visit
% http://creativecommons.org/licenses/by/4.0/.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[12pt]{ijck}

%\usepackage{listings}

\usepackage{fancyvrb,newverbs,xcolor}

\definecolor{cverbbg}{gray}{0.93}
\colorlet{pyyellow}{yellow!10!white}
\colorlet{yamlblack}{black!10!white}

% \newcommand{\yabox}[1]{\colorbox{yamlblack}{\texttt{#1}}}
% \newcommand{\pybox}[1]{\colorbox{pyyellow}{\texttt{#1}}}
\newverbcommand{\yabox}
  {\setbox\verbbox\hbox\bgroup}
  {\egroup\colorbox{yamlblack}{\box\verbbox}}
\newverbcommand{\pybox}
  {\setbox\verbbox\hbox\bgroup}
  {\egroup\colorbox{pyyellow}{\box\verbbox}}

@@

This section will provide a brief introduction to the \ck{} database format.
This is not a complete listing of available fields and more complete examples of
\ck{} files can be found in the \pybox|tests| directory of the \pk{} package. In
addition, in Section~\ref{sec:usage-example}, we will provide examples of \ck{}
files. We indicate \yabox|yaml| keywords or values using text with gray background,
and \pybox|Python| code using text with a yellow background.

@@

The \yabox|value| can be a string, integer, or floating point number. The \ck{}
format is designed to include all of the information necessary to simulate a
given experiment and can be broken into two main sections. The first section
details all of the ``meta'' information about the experiment and the \ck{} file.
In this section, the common \yabox|author| mapping is used for any fields where
an author must be specified. The \yabox|author| mapping contains the following
fields:
%
\begin{itemize}
    \item \yabox|name| (required, string): The author's full name
    \item \yabox|ORCID| (optional, string): the author's ORCID number
\end{itemize}
%
More generally for the meta information, the required keys in this section
are:
%
\begin{itemize}
    \item \yabox|file-author| (required, \yabox|author|-type mapping): The
    author of the \ck{} file
    \item \yabox|file-version| (required, integer): The version of the \ck{}
    file
    \item \yabox|chemked-version| (required, string): The version of the \ck{}
    schema that this file conforms to
    \item \yabox|reference| (required, mapping): The reference information for
    the article associated with the data in the file. The fields in this
    mapping are:
    \begin{itemize}
        \item \yabox|authors| (required, sequence): A sequence of \yabox|author|
        mappings
        \item \yabox|journal| (required, string): The name of the publishing
        journal
        \item \yabox|year| (required, integer): The year of publication
        \item \yabox|doi| (optional, string): The article DOI
        \item \yabox|volume| (required, integer): The journal volume number
        \item \yabox|pages| (optional, string): The article pages
        \item \yabox|detail| (optional, string): A description of from where the
        data originated (e.g., figure or table number)
    \end{itemize}
    \item \yabox|experiment-type| (required, string): Currently, only
    \yabox|ignition delay| is supported
    \item \yabox|apparatus| (required, string): The type of apparatus used to
    perform the experiment (\yabox|shock tube| or \yabox|rapid compression machine|)
\end{itemize}

The second section of the file encodes the experimental data. In the \ck{} file,
this is done in a sequence of mappings, with the following information required
in each element of the sequence:
%
\begin{itemize}
    \item \yabox|temperature| (required, sequence): The temperature of the
    experiment, with units
    \item \yabox|ignition-delay| (required, sequence): The ignition delay of the
    experiment, with units
    \item \yabox|pressure| (required, string): The pressure of the experiment,
    with units
    \item \yabox|composition| (required, sequence): The composition of the
    mixture in the experiment, a sequence of mappings with the following fields
    \begin{itemize}
        \item \yabox|species-name| (required, string): The name of the species
        \item \yabox|mole-fraction|, \yabox|mass-fraction|, \yabox|mole-percent|
        (required, float): The mole fraction or percent, or the mass fraction.
        Only one of the three is permitted in a given \yabox|composition|
        sequence.
        \item \yabox|InChI|, \yabox|SMILES|, \yabox|atomic-composition|, or
        \yabox|elemental-composition| (required, string or sequence): The InChI
        or SMILES representing the molecule, or its atomic or elemental
        composition (\yabox|atomic-composition| and
        \yabox|elemental-composition| are synonyms) as a sequence
    \end{itemize}
    \item \yabox|ignition-type| (required, mapping): The type of ignition
    measured in the experiments. The required fields are
    \begin{itemize}
        \item \yabox|type| (required, string): How the ignition delay was
        measured; one of \yabox|d/dt max| to indicate the ignition point is
        found at the maximum of the time derivative of the \yabox|target|,
        \yabox|max| or \yabox|min| to indicate the ignition point is at the
        maximum or minimum of the \yabox|target|, or \yabox|1/2 max| to
        indicate the half-maximum point
        \item \yabox|target| (required, string): The target for the type of the
        ignition point; one of \yabox|temperature|, \yabox|pressure|,
        \yabox|OH|, \yabox|OH*|, \yabox|CH|, or \yabox|CH*|
    \end{itemize}
\end{itemize}

The top-level key for the sequence is called \yabox|datapoints|. Each of the
quantities in an element must be specified with units. This should be done as a
single string associated with the first element of the sequence. The units of
the quantity are validated to ensure they have the appropriate dimensionality
for the quantity. In addition, each of the quantities in an element can
optionally be assigned an uncertainty. This uncertainty can be either absolute
or relative, and is specified as elements of the sequence of the associated key.
For example, the absolute uncertainty of the temperature and the relative
uncertainty of the ignition delay might be specified as follows:

@@

The basic user interface to the \ck{} file in \pk{} is through the
\pybox|ChemKED| class. The \pybox|ChemKED| class constructor takes the filename
of a \ck{} file or a Python dictionary containing the contents of a \ck{} file
as its argument. When the file or dictionary is loaded, it is validated to have
the correct format by the Python package Cerberus \autocite{cerberus}. The
schema used for validation of the \ck{} files is available publicly in the
source code repository for
PyKED\footnote{\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/chemked_schema.yaml}}.

The fields of the \ck{} file are stored as instance attributes of the
\pybox|ChemKED| class. The following attributes are available:
%
\begin{itemize}
    \item \pybox|chemked_version|, \pybox|file_version|, \pybox|file_author|,
    \pybox|experiment_type|: Stores the values from the \ck{} file
    \item \pybox|reference|: Instance of a \pybox|namedtuple| containing all
    the information from the literature reference associated with the data
    \item \pybox|apparatus|: Instance of a \pybox|namedtuple| containing all
    the information about the apparatus used to perform the experiment
    \item \pybox|datapoints|: A Python list of \pybox|DataPoint| instances
\end{itemize}

The \pybox|DataPoint| class stores the information associated with a single data
point in the \ck{} file (i.e., a single element of the \yabox|datapoints|
sequence). Similar to the \pybox|ChemKED| class, information in the
\pybox|DataPoint| is stored as instance attributes:
%
\begin{itemize}
    \item \pybox|equivalence_ratio|: The value of the equivalence ratio, if
    present. For informational purposes only, no validation is done of the
    value.
    \item \pybox|ignition_delay|, \pybox|temperature|, \pybox|pressure|,
    \pybox|pressure_rise|: The value of these data are stored as instances of
    the \pybox|Quantity| class from the Pint \autocite{Grecco2016} package,
    possibly with associated uncertainty
    \item \pybox|composition|: A list of dictionaries of the species and their
    respective amounts. The values are validated so that \yabox|mole-percent|,
    \yabox|mole-fraction|, or \yabox|mass-fraction| cannot be mixed for a single
    data point, and so that the sum of the values is approximately 1.0, or
    100.0 for \yabox|mole-percent|.
    \item \pybox|composition_type|: A string indicating the type of composition
    information for the data point---one of \pybox|'mole-percent'|,
    \pybox|'mole-fraction'|, or \pybox|'mass-fraction'|
    \item \pybox|volume_history|: If the \yabox|volume-history| of an RCM
    experiment is provided in the \ck{} file, it is stored in this attribute as
    a \pybox|namedtuple|, and the actual values are stored in NumPy arrays
    \autocite{vanderWalt:2011np}
    \item \pybox|compression_time|: If the \ck{} file encodes an RCM experiment,
    the length of the compression stroke may be included
    \item \pybox|ignition_type|: A dictionary specifying the method of the
    measurement of ignition delay
\end{itemize}

The \pybox|DataPoint| class defines two instance methods:
\pybox|get_cantera_mole_fraction()| and \pybox|get_cantera_mass_fraction()|.
With these methods, the composition of the reactant mixture can be output to a
format that can be used to set the composition of a
Cantera~\pybox|Solution|~\autocite{Cantera:2.3.0}.
The \pybox|composition| specification does not contain the molecular
weights of the components, so conversion between mole fractions and mass
fractions is not currently possible.

The \pybox|ChemKED| class defines one instance method:
\pybox|get_dataframe()|. Using this method, the user can get an instance of a
Pandas \pybox|DataFrame|~\autocite{pandas} that contains the information in the
list of \pybox|DataPoint|s. The user can specify the columns included in the
\pybox|DataFrame| by passing a list of column names to the
\pybox|output_columns| argument of the \pybox|get_dataframe()| method. The
possible columns are not case-sensitive and are:

\noindent\begin{itemize*}
    \item \pybox|'Temperature'|
    \item \pybox|'Pressure'|
    \item \pybox|'Ignition Delay'|
    \item \pybox|'Composition'|
    \item \pybox|'Equivalence Ratio'|
    \item \pybox|'Reference'|
    \item \pybox|'Apparatus'|
    \item \pybox|'Experiment Type'|
    \item \pybox|'File Author'|
    \item \pybox|'File Version'|
    \item \pybox|'ChemKED Version'|
\end{itemize*}

In addition, specific fields from the \pybox|Reference| and \pybox|Apparatus|
attributes can be included by specifying the name after a colon. These options
are:

\noindent\begin{itemize*}
    \item \pybox|'Reference:Volume'|
    \item \pybox|'Reference:Journal'|
    \item \pybox|'Reference:DOI'|
    \item \pybox|'Reference:Authors'|
    \item \pybox|'Reference:Detail'|
    \item \pybox|'Reference:Year'|
    \item \pybox|'Reference:Pages'|
    \item \pybox|'Apparatus:Kind'|
\end{itemize*}

% This weirdness is necessary to prevent the list from running over the margin
\noindent\begin{itemize*}
    \item \pybox|'Apparatus:Facility'|
    \item \pybox|'Apparatus:Institution'|
\end{itemize*}

Only the first author is printed when \pybox|Reference| or
\pybox|Reference:Authors| is selected because the whole author list may be
quite long.

@@

\pk{} is licensed under the permissive, open-source BSD 3-clause license. The
source code is publicly available on GitHub at
\url{https://github.com/pr-omethe-us/PyKED}~\autocite{PyKED}. Installation of stable versions
can be accomplished from a package repository, using either \pybox|pip| to install
from \href{https://pypi.python.org/pypi}{PyPI} (Python Package Index), or
\pybox|conda| to install from the
\href{https://anaconda.org/conda-forge}{\texttt{conda-forge}} channel. In
addition, \pk{} can be installed from the source code by running
\pybox|python setup.py install| or \pybox|pip install .| from the root
directory of the repository.

@@

The \ck{} file that will be used in this example can be found in the
\pybox|tests| directory of the PyKED
repository\footnote{\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/tests/testfile_rcm.yaml}}.
Examining that file, we find the first section specifies the information about
the \ck{} file itself:
%
\begin{yamlbox}
file-author:
  name: Kyle E Niemeyer
  ORCID: 0000-0003-4425-7097
file-version: (1, 0)
chemked-version: 0.0.1
\end{yamlbox}
%
Then, we find the information regarding the article in the literature from which
this data was taken. In this case, the dataset comes from the work of
\textcite{Mittal2006a}:
%
\begin{yamlbox}
reference:
  doi: 10.1002/kin.20180
  authors:
    - name: Gaurav Mittal
    - name: Chih-Jen Sung
      ORCID: 0000-0003-2046-8076
    - name: Richard A Yetter
  journal: International Journal of Chemical Kinetics
  year: 2006
  volume: 38
  pages: 516-529
  detail: Fig. 6, open circle
experiment-type: ignition delay
apparatus:
  kind: rapid compression machine
  institution: Case Western Reserve University
  facility: CWRU RCM
\end{yamlbox}
%
Finally, there is just a single datapoint in this file, which describes the ignition
delay for the experiment, the mixture composition, the initial temperature,
pressure, compression time, ignition type, and the volume history that specifies
how the volume of the reactor varies with time, for simulating the compression
stroke and post-compression processes.
%
\begin{yamlbox}
datapoints:
  - temperature: 297.4 kelvin
    ignition-delay: 1.0 ms
    pressure: 958.0 torr
    composition:
      - species-name: H2
        InChI: 1S/H2/h1H
        mole-fraction: 0.12500
      - species-name: O2
        InChI: 1S/O2/c1-2
        mole-fraction: 0.06250
      - species-name: N2
        InChI: 1S/N2/c1-2
        mole-fraction: 0.18125
      - species-name: Ar
        InChI: 1S/Ar
        mole-fraction: 0.63125
    ignition-type:
      target: pressure
      type: d/dt max
    compression-time: 38.0 ms
    volume-history:
      time:
        units: s
        column: 0
      volume:
        units: cm3
        column: 1
      values:
        - [0.00E+000, 5.47669375000E+002]
        - [1.00E-003, 5.46608789894E+002]
        - [2.00E-003, 5.43427034574E+002]
        - [3.00E-003, 5.38124109043E+002]
        ...
\end{yamlbox}
%
The values for the \yabox|volume-history| are truncated here to save space. A user
might want to load the information from this file to perform a simulation using
Cantera to calculate the ignition delay. All of the information required to
perform this simulation is present in the \ck{} file, with the exception of a
chemical kinetic model for \ce{H2}\slash {CO} combustion.
%
\begin{pythonbox}
import cantera as ct
from pyked import ChemKED

# Load the ChemKED file and retrieve the first element of the
# datapoints list, which is an instance of the DataPoint class
ck = ChemKED('testfile_rcm.yaml')
dp = ck.datapoints[0]
\end{pythonbox}
%
The initial temperature, pressure, and mixture composition can be read from the
instance of the \pybox|DataPoint| class. \pk{} uses Pint \pybox|Quantities| to
store values with units, while Cantera expects a floating point value in SI
units as input. Therefore, we use the built-in capabilities of Pint to convert
the units from those specified in the \ck{} file to SI units:
%
\begin{pythonbox}
T_initial = dp.temperature.to('K').magnitude
P_initial = dp.pressure.to('Pa').magnitude
X_initial = dp.get_cantera_mole_fraction()

@@

# Create the reactor and the outside environment
reac = ct.IdealGasReactor(gas)
env = ct.Reservoir(ct.Solution('air.xml'))
\end{pythonbox}
%
To apply the effect of the volume trace to the \pybox|IdealGasReactor|, a
\pybox|Wall| is installed between the reactor and the environment and assigned
a velocity. Although we do not show the details here, a reference implementation
of a class that computes a wall velocity given the volume history of the reactor
is available in CanSen~\autocite{cansen}, in the
\pybox|cansen.profile.VolumeProfile| class. Then, a Cantera \pybox|ReactorNet|
can be used to advance the state through autoignition, in this case to an end
time of \SI{50}{\milli\second}:
%
\begin{pythonbox}
time = dp.volume_history.time
volume = dp.volume_history.volume
ct.Wall(reac, env, velocity=VolumeProfile(time=time, volume=volume))

@@

# Integrate for 50 ms
while netw.time < 0.05:
    netw.step()
\end{pythonbox}
%
Although not shown in this example, the user would typically store information
about the \pybox|IdealGasReactor| (e.g., temperature, pressure, mass fractions)
at each time step of the integration for post-processing.

@@

The \ck{} file used in this example can be found in the \pybox|tests| directory of the \pk{}
repository\footnote{\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/tests/testfile_st_p5.yaml}}.
The data in this file comes from Stranic et al.~\autocite{Stranic:2012}, describing
shock-tube ignition delays for \textit{tert}-butanol. We have omitted the file
meta information below for space; the format is largely similar to the example in Section~\ref{sec:rcm-modeling}.
%
This \ck{} file specifies multiple data points with some common
conditions, including a common mixture composition and common definition of
ignition delay. Therefore, a \yabox|common-properties| section is specified.
%
\begin{yamlbox}
common-properties:
  composition: &comp
    - species-name: t-butanol
      InChI: 1S/C4H10O/c1-4(2,3)5/h5H,1-3H3
      mole-fraction: 0.003333333
    - species-name: O2
      InChI:  1S/O2/c1-2
      mole-fraction: 0.04
    - species-name: Ar
      InChI:  1S/Ar
      mole-fraction: 0.956666667
  ignition-type:  &ign
    target: OH* concentration
    type: 1/2 max
\end{yamlbox}
%
This block uses the ability of YAML files to define an anchor with the
\yabox|&| symbol and refer to that section later with the \yabox|*| symbol, as
can be seen in the definition of the \yabox|datapoints| section:
%
\begin{yamlbox}
datapoints:
  - temperature: 1459 kelvin
    ignition-delay: 347 us
    pressure: 1.60 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1389 kelvin
    ignition-delay: 756 us
    pressure: 1.67 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1497 kelvin
    ignition-delay: 212 us
    pressure: 1.55 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1562 kelvin
    ignition-delay: 105 us
    pressure: 1.50 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
\end{yamlbox}
The \yabox|common-properties| section is not required, but can save space and
help avoid errors when all data points share some common values.
